### 2022-02-15 introduction/introduction.tex
more

Half sleeping, half playing Fruit Ninja on my smartphone at one of the
philosophy lectures in my high school I heard my professor exclaim with
excitement: “Mathematical truth is the highest and absolute form of truth!”.
Unbothered and uninterested, I continued playing Fruit Ninja.

@@

The proofs need to be spelled out in minute details to be accepted by a proof assistant.
Even though many proof assistants offer some sort of automation, seemingly simple
statements cannot be automatically proved. This is one of the reasons why large
verification projects can take years to finish.

@@

This approach proved useful: around 69\% of the simple proof goals can be
discharged automatically using Sledgehammer \cite{kb-13-mash}. However, there is
clearly some untapped potential in the approach as it communicates with
automatic theorem provers through a translation. To bridge this translation gap,
attempts were made to use higher-order automatic theorem provers instead of
first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
higher-order theorem provers of the time were primarily designed for small and
tricky problems. On the other hand, problems coming from hammers consist of
mostly first-order formulas, are rather large, and mostly require simple
higher-order proof steps. Thus, from a standpoint of proof assistant ideal
automatic theorem prover fulfills the following wishlist:
\begin{enumerate}
  \item\label{it:fo} Performs as best first-order provers on first-order problems
  \item\label{it:ho} Supports higher-order logic and scales with the amount of higher-order axioms
  \item\label{it:size} Scales well with the size of the problem
\end{enumerate}
This brings us to the topic of this thesis: \emph{How can we develop a theorem provers that fulfills this wishlist?}

\paragraph{Our Approach} To improve the automatic reasoning for higher-order
logic we decided not to start from a blank slate and build a new theorem prover.
Rather, we start from a position of strength --- from state-of-the-art
first-order superposition theorem prover E \cite{scv-19-e23} --- and extend this
prover to higher-order logic. In this way, the points \ref{it:fo} and
\ref{it:size} from the above wishlist  is fulfilled for free, as E is very
efficient and deals well with large problems.

We do not extend E to higher-order logic in one atomic step. Instead, we extend
it by adding support for three distinguished features of higher-order logic,
which gradually increase its support for higher-order logic. With the layered
approach we make sure that after every extension to higher-order logic, E's
efficiency on first-order problems was not affect. In other words, while trying
to fulfill the point \ref{it:ho} from the wishlist we make sure that points
\ref{it:fo} and \ref{it:size} are not affected.

We have split the road to full higher-order logic in three stages. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\forall x. x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with addition function ($\cst{+}$) is enough to
disprove this statement. The first stage is to extend E to support
quantification over functions that are already present in the theory.

Higher-order logic also supports expressing functions that are not explicitly
stated in the problem. For example, suppose we want to prove that there is a
function such that returns 4 when given 2 and 5 when given 3 ($\exists x.
x(\cst{2}) \eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such
function is the function that adds 2 to its argument. In higher-order logic such
unnamed function can be represented using $\lambda$ abstractions as $\lambda
x.\, x \, \cst{+} \, 2$. Replacing $x$ with this term suffices to prove the
original statement. The second stage is to support automatically synthesizing
such functions.

In higher-order logic we can make statements about statements. For example, we
can naturally represent statements such as ``whatever Jasmin says is true'' as
$\forall x. \, \cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support this
feature of first-order logic.

>>>>

Half sleeping, half playing Fruit Ninja on my smartphone at one of the
philosophy lectures in my high school I heard my professor exclaim with
excitement: “Mathematical truth is the highest form of truth!”.
Unbothered and uninterested, I continued playing Fruit Ninja.

@@

The proofs need to be spelled out in minute detail to be accepted by a proof assistant.
Even though many proof assistants offer some sort of automation, seemingly simple
statements cannot be automatically proved. This is one of the reasons why large
verification projects can take years to finish.

@@

This approach proved useful: around 76\% of the simple proof goals can be
discharged automatically using Sledgehammer \cite{bgkku-16-larning-fact-selector}. However, there is
clearly some untapped potential in the approach as it communicates with
automatic theorem provers through a translation. To bridge this translation gap,
attempts were made to use higher-order automatic theorem provers instead of
first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
higher-order theorem provers of the time were primarily designed for small and
tricky problems. On the other hand, problems coming from hammers consist of
mostly first-order formulas, are rather large, and mostly require simple
higher-order proof steps. Thus, from a standpoint of proof assistant ideal
automatic theorem prover fulfills the following wishlist:
\begin{enumerate}
  \item\label{it:fo} Performs as best first-order provers on first-order problems
  \item\label{it:ho} Supports higher-order logic and scales with the amount of higher-order axioms
  \item\label{it:size} Scales well with the size of the problem
\end{enumerate}
This brings us to the topic of this thesis: \emph{How can we develop a theorem provers that fulfills this wishlist?}

\paragraph{Our Approach} To improve the automatic reasoning for higher-order
logic we decided not to start from a blank slate and build a new theorem prover.
Rather, we start from a position of strength --- from state-of-the-art
first-order superposition theorem prover E \cite{scv-19-e23} --- and extend this
prover to higher-order logic. In this way, the points \ref{it:fo} and
\ref{it:size} from the above wishlist are fulfilled for free, as E is very
efficient and deals well with large problems.

We do not extend E to higher-order logic in one atomic step. Instead, we extend
it by adding support for three distinguished features of higher-order logic,
which gradually increase its support for higher-order logic. With the layered
approach we make sure that after every extension to higher-order logic, E's
efficiency on first-order problems did not decrease. In other words, while trying
to fulfill the point \ref{it:ho} from the wishlist we make sure that points
\ref{it:fo} and \ref{it:size} are not affected.

We have split the road to full higher-order logic in three stages. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\forall x. \,  x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with addition function ($\cst{+}$) is enough to
disprove this statement. The first stage is to extend E to support
quantification over functions that are already present in the theory.

Higher-order logic also supports expressing functions that are not explicitly
present in the theory. For example, suppose we want to prove that there is a
function such that returns 4 when given 2 and 5 when given 3 ($\exists x.
x(\cst{2}) \eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such
function is the function that adds 2 to its argument. In higher-order logic such
unnamed function can be represented using $\lambda$ abstractions as $\lambda
x.\, x \, \cst{+} \, 2$. Replacing $x$ with this term suffices to prove the
original statement. The second stage is to support automatically synthesizing
such functions.

In higher-order logic we can make statements about statements. For example, we
can naturally represent statements such as ``what Jasmin says is true'' as
$\forall x. \, \cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support this
feature of first-order logic.

\section{Contribution}
\section{Publications and Thesis Structure}
\section{Related Work}
